Global patterns of forest autotrophic carbon fluxes

### 2020-02-17 manuscript/manuscript draft.Rmd
updated hypothesis table and corrected citations in manuscript draft

**FACF can be influenced by many other factors as well**, which often act across a range of scales, and may show interactive effects with each other [@cleveland_relationships_2011]. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], biodiversity [@liang_positive_2016], management [@simova_enigma_2017]; nutrient availability [@aragao_above-_2009]; and altitude [@girardin_net_2010; @malhi_variation_2017] all impact FACF. On a global scale, we expect that FACF are most strongly influenced by broad climatic gradients. There is evidence that FACFs also respond to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005] in potentially significant ways. Furthermore, MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and FACFs is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between primary productivity and MAT, suggesting that the effect of temperature is due to increased length of growing season, rather than an inherent influence of temperature on FACF.

**The recent development of a global forest carbon database synthesizing multiple variables and including records of stand history (ForC; [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc:_2018]) opens up the possibility for a standardized analysis of global scale variation** in multiple FACF and the principle climatic drivers of these patterns. In order to approach this broad topic, we simplify the major gaps in our knowledge to five broad questions and corresponding hypotheses (Table 1). ** see [issue #48 ](https://github.com/forc-db/Global_Productivity/issues/48)** First, we ask how forest autotrophic carbon fluxes (FACF) vary with latitude. We then test how these fluxes relate to MAT and MAP, and additionally how they respond to other, less well-studied, climate variables. Finally, we consider the relationship between FACF and seasonality, considering the role of seasonality in explaining variation in carbon fluxes, and the influence of climate on FACF standardized by growing season length. We use a comprehensive global database of forest carbon fluxes to address the above questions for nine carbon fluxes, allowing for an in-depth exploration of the effect of climate on FACF globally.

@@

Analyses were conducted on data contained in the open-access ForC database [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc:_2018]. This database contains records of field-based measurements of forest carbon stocks and annual fluxes, compiled from original publications and existing data compilations and databases. Associated data, such as stand age, measurement methodologies, and disturbance history, are also included. The database was significantly expanded since the publication of [@anderson-teixeira_forc:_2018] through integration with the Global Soil Respiration Database [@bond-lamberty_global_2010]. Additional targeted literature searches were conducted to identify any further available data on FACF, with particular focus on mature forests in temperate and boreal regions. ForC currently contains 29730 records from 4979 plots, representing 20 distinct ecozones across all forested biogeographic and climate zones. We used ForC v3.0, archived on Zenodo with DOI 10.5281/zenodo.3403855.

@@

A subset of the ForC database was generated for the purposes of this analysis, in order to control for data quality and remove biasing factors. Since management can alter observed patterns of FACF [@simova_enigma_2017], sites were excluded from analysis if they were managed, defined as plots that were planted, managed as plantations, irrigated, fertilised or including the term "managed" in their site description. Sites that had experienced significant disturbance were also excluded. Disturbances that justified site exclusion were major cutting or harvesting, and/or burning, flooding, drought and storm events with site mortality >10% of trees. Grazed sites were retained.  
There is evidence that stand age influences patterns of FACF and carbon allocation in forest ecosystems, and can confound relationships between latitude and primary productivity [@de_lucia_forest_2007; @gillman_latitude_2015]. To reduce any biasing effects of stand age, we included only stands of known age $\ge$ 100 years and those described by terms such as "mature", "intact", or "old-growth".

@@

*Climate datasets.* Where site-level data on mean annual temperature, mean annual precipitation, and latitude were available in the primary literature, this data was compiled and entered directly into the ForC database. Based on the geographic co-ordinates for each site, data on a further 11 climate variables was extracted from five open-access climate datasets: WorldClim [@hijmans_very_2005], WorldClim2 [@fick_worldclim_2017], the Climate Research Unit (CRU) time-series dataset v. 4.03 [@harris_updated_2014], the Global Aridity Index and Potential Evapotranspiration Climate Database [@trabucco_global_2018], and TerraClimate [@abatzoglou_terraclimate_2018] (see Supplementary Information S1 for details of climate variables). Where site-level data was missing for mean annual temperature and/or mean annual precipitation, data was extracted from the WorldClim dataset.

@@

Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*; **REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship is approximately linear (Fig. 2). This contrasts with suggestions that productivity of temperate forests is similar to that of tropical forests (**REFS**). Temperate forests tend to be younger than tropical forests (**REF**), so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects ([Poulter et al. 2018-DOI:10.1594/PANGAEA.889943](http://pure.iiasa.ac.at/id/eprint/15361/); Potapov et al. 2008. Mapping the World’s Intact Forest Landscapes by Remote Sensing. Ecology and Society 13 (2), 51. ). In addition, because C allocation varies with stand age [@de_lucia_forest_2007] (**See Nobby's comment in manuscript-draft_NK.pdf**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (** Nobby AFM paper: Understanding the controls over forest carbon use efficiency on small spatial scales: Effects of forest disturbance and tree diversity**) [*purpose for respiration/ other compenents (**See Nobby's comment in manuscript-draft_NK.pdf**)].

@@

bbl: remove this sentence? Of particular interest and significance are the relationships amongst $GPP$, net primary productivity ($NPP$ and its components, particularly $ANPP_{woody-stem}$), and respiration ($R_{auto}$ and components). There have been suggestions that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which are based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@de_lucia_forest_2007; [Malhi 2012](https://www.researchgate.net/publication/227837530_The_productivity_metabolism_and_carbon_cycle_of_tropical_forest_vegetation); @anderson-teixeira_carbon_2016], but our analysis suggests that these low values might more appropriately be attributed to the fact that these forests are old than to their tropical climate. Indeed, $CUE$ is known to decline with forest age ([@de_lucia_forest_2007]; [Collalti & Prentice 2019](https://academic.oup.com/treephys/article-abstract/39/8/1473/5423351?redirectedFrom=fulltext)) but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. Among our sites with relevant data, there is indication that CUE or $ANPP_{woody-stem}$/$GPP$ increase with latitude (**some SI table**). Additional measurements with careful methodological standardization across a consistent set of mature forest sites will be necessary to determine whether any climate-driven gradients in allocation exist at the global scale.

>>>>

**FACF can be influenced by many other factors as well**, which often act across a range of scales, and may show interactive effects with each other [@cleveland_relationships_2011]. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], biodiversity [@liang_positive_2016], management [@simova_enigma_2017]; nutrient availability [@aragao_above_2009]; and altitude [@girardin_net_2010; @malhi_variation_2017] all impact FACF. On a global scale, we expect that FACF are most strongly influenced by broad climatic gradients. There is evidence that FACFs also respond to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005] in potentially significant ways. Furthermore, MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and FACFs is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between primary productivity and MAT, suggesting that the effect of temperature is due to increased length of growing season, rather than an inherent influence of temperature on FACF.

**The recent development of a global forest carbon database synthesizing multiple variables and including records of stand history (ForC; [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018]) opens up the possibility for a standardized analysis of global scale variation** in multiple FACF and the principle climatic drivers of these patterns. In order to approach this broad topic, we simplify the major gaps in our knowledge to five broad questions and corresponding hypotheses (Table 1). ** see [issue #48 ](https://github.com/forc-db/Global_Productivity/issues/48)** First, we ask how forest autotrophic carbon fluxes (FACF) vary with latitude. We then test how these fluxes relate to MAT and MAP, and additionally how they respond to other, less well-studied, climate variables. Finally, we consider the relationship between FACF and seasonality, considering the role of seasonality in explaining variation in carbon fluxes, and the influence of climate on FACF standardized by growing season length. We use a comprehensive global database of forest carbon fluxes to address the above questions for nine carbon fluxes, allowing for an in-depth exploration of the effect of climate on FACF globally.

@@

Analyses were conducted on data contained in the open-access ForC database [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018]. This database contains records of field-based measurements of forest carbon stocks and annual fluxes, compiled from original publications and existing data compilations and databases. Associated data, such as stand age, measurement methodologies, and disturbance history, are also included. The database was significantly expanded since the publication of [@anderson-teixeira_forc_2018] through integration with the Global Soil Respiration Database [@bond-lamberty_global_2010]. Additional targeted literature searches were conducted to identify any further available data on FACF, with particular focus on mature forests in temperate and boreal regions. ForC currently contains 29730 records from 4979 plots, representing 20 distinct ecozones across all forested biogeographic and climate zones. We used ForC v3.0, archived on Zenodo with DOI 10.5281/zenodo.3403855.

@@

A subset of the ForC database was generated for the purposes of this analysis, in order to control for data quality and remove biasing factors. Since management can alter observed patterns of FACF [@simova_enigma_2017], sites were excluded from analysis if they were managed, defined as plots that were planted, managed as plantations, irrigated, fertilised or including the term "managed" in their site description. Sites that had experienced significant disturbance were also excluded. Disturbances that justified site exclusion were major cutting or harvesting, and/or burning, flooding, drought and storm events with site mortality >10% of trees. Grazed sites were retained.  
There is evidence that stand age influences patterns of FACF and carbon allocation in forest ecosystems, and can confound relationships between latitude and primary productivity [@delucia_forest_2007; @gillman_latitude_2015]. To reduce any biasing effects of stand age, we included only stands of known age $\ge$ 100 years and those described by terms such as "mature", "intact", or "old-growth".

@@

*Climate datasets.* Where site-level data on mean annual temperature, mean annual precipitation, and latitude were available in the primary literature, this data was compiled and entered directly into the ForC database. Based on the geographic co-ordinates for each site, data on a further 11 climate variables was extracted from five open-access climate datasets: WorldClim [@hijmans_very_2005], WorldClim2 [@fick_worldclim_2017], the Climate Research Unit (CRU) time-series dataset v. 4.03 [@harris_updated_2014], the Global Aridity Index and Potential Evapotranspiration Climate Database [@trabucco_global_2019], and TerraClimate [@abatzoglou_terraclimate_2018] (see Supplementary Information S1 for details of climate variables). Where site-level data was missing for mean annual temperature and/or mean annual precipitation, data was extracted from the WorldClim dataset.

@@

Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*; **REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship is approximately linear (Fig. 2). This contrasts with suggestions that productivity of temperate forests is similar to that of tropical forests (**REFS**). Temperate forests tend to be younger than tropical forests (**REF**), so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects ([Poulter et al. 2018-DOI:10.1594/PANGAEA.889943](http://pure.iiasa.ac.at/id/eprint/15361/); Potapov et al. 2008. Mapping the World’s Intact Forest Landscapes by Remote Sensing. Ecology and Society 13 (2), 51. ). In addition, because C allocation varies with stand age [@delucia_forest_2007] (**See Nobby's comment in manuscript-draft_NK.pdf**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (** Nobby AFM paper: Understanding the controls over forest carbon use efficiency on small spatial scales: Effects of forest disturbance and tree diversity**) [*purpose for respiration/ other compenents (**See Nobby's comment in manuscript-draft_NK.pdf**)].

@@

bbl: remove this sentence? Of particular interest and significance are the relationships amongst $GPP$, net primary productivity ($NPP$ and its components, particularly $ANPP_{woody-stem}$), and respiration ($R_{auto}$ and components). There have been suggestions that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which are based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@delucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016], but our analysis suggests that these low values might more appropriately be attributed to the fact that these forests are old than to their tropical climate. Indeed, $CUE$ is known to decline with forest age [@delucia_forest_2007; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. Among our sites with relevant data, there is indication that CUE or $ANPP_{woody-stem}$/$GPP$ increase with latitude (**some SI table**). Additional measurements with careful methodological standardization across a consistent set of mature forest sites will be necessary to determine whether any climate-driven gradients in allocation exist at the global scale.
